---
title: "Pilot 2: Exploring biodiversity impacts of oil-palm plantations in Indonesia and Malaysia"
author: "Adrienne Etard (International Institute for Applied Systems Analysis); & Rui Song (Global Canopy); Claudia Gray (Global Canopy); Nicola Wilson (Global Canopy); Fiona Pedeboy (Global Canopy); Gerado Lopez Saldana (Assimila)"
date: last-modified
date-format: long
format: html
editor: visual
editor_options: 
  chunk_output_type: console
htlm:
    toc: true
    toc-depth: 3
    toc-location: left
---

# Aims

This work is part of the Pilot 2 (Agrifood Supply Chains) of the LEON project (Leveraging Earth Observation for Nature Finance), funded by ESA ([project website here](https://www.leon-naturefinance.org/)).

The aims of this work are to:

-   1: Derive different biodiversity indices (average abundance, similarity, and the Biodiversity Intactness Index - BII) for oil-palm plantations focusing on Pilot 2's focal area, Indonesia and Malaysia.

-   2: Explore the role and ability of Earth Observation data in explaining and predicting biodiversity change in the pilot region.

-   3: Derive BII for the mill areas of interest, as a useful indicator for financial institutions.

The approach is based on the PREDICTS database [Hudson et al. 2014](https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.1303), which is a collection of independent studies that have sampled biodiversity *in-situ*, in different land-use types. PREDICTS notably underpins the global calculations of the BII. In PREDICTS-based approaches, biodiversity change is inferred from a space-for-time substitution, by comparing assumed pristine locations (sites sampled in minimally-used primary vegetation) to disturbed sites (e.g. located in cropland, secondary vegetation, or plantations).

Here, we specifically investigates whether we can improve BII predictions by considering EO-derived data in the biodiversity models that underpin BII (using EO-derived data as explanatory variables for biodivrsity change). In particular, we consider the following EO-derived datasets:

-   Oil-palm plantation age. We investigate wheter plantation age influences the biodiversity indicators (abundance). For plantations at a global scale, [Tudge et al. (2023)](https://www.sciencedirect.com/science/article/pii/S2351989423003864) found a significant effect for species richness. Note that Tudge et al. (2023) did not detect any effect of age on abundance in oil palm plantations.

Here, plantation age is derived from [Danylo et al. (2021)](https://www.nature.com/articles/s41597-021-00867-1), who produced a map of the extent and age of detection of oil palm plantations encompassing Malaysia and Indonesia.

This document thus presents the step-by-step analyses (including loading data, code, and deriving outputs) for investigating the biodiversity impacts of oil-palm plantations in Indonesia and Malaysia using the PREDICTS database.

Acknowledgements: Adriana de Palma wrote a [tutorial](https://adrianadepalma.github.io/BII_tutorial/#Run_the_statistical_analysis) for deriving BII using the PREDICTS database, which was a useful source in developping this work. Tim Newbold wrote [useful packages](https://github.com/timnewbold/StatisticalModels) for data preparation and model fitting that we employed in this work.

```{r}
#| output: false ## this is to control whether outputs are rendered
#| echo: true ## this is to control whether code is rendered
#| label: start time
TimeStart <- Sys.time()
```

# 1. DATA PREPARATION

Load the necessary packages; load and prepare the PREDICTS data and otherwise EO-derived layers that we employ. The PREDICTS data can be downloaded from [here](https://data.nhm.ac.uk/dataset/?tags=predicts).

## 1.1. Load packages and useful functions

```{r}
#| output: false
#| echo: true
Packages <- c(
  'dplyr',
  'ggplot2',
  'rnaturalearth',
  'StatisticalModels',
  'ggpubr',
  'purrr',
  'betapart',
  'vegan',
  'terra',
  'sp'
)
sapply(Packages, FUN=library, character.only=TRUE)
rm(Packages)

source('Source_Functions.R')
```

## 1.2. Load the land-cover layers for the pilot region (for projections)

In the work conducted for Pilot 2, Rui Song prepared land cover maps for our study region, with corrections for the detection of oil-palm plantations. We use these maps to predict BII around mills.

However, we need to get some information on the state of forests. Indeed, tree cover alone is not enough as it does not reflect possible forest degradation. To this end, we overlay a dataset on forest integrity (the Forest Landscape Integirty Index; https://www.forestintegrity.com/home)

```{r}
#| output: false 
#| echo: true 
#| label: loading and preparing land cover layers for Indonesia and Malaysia

Indonesia_Malaysia_Map <- vect(rnaturalearth::ne_countries(scale = "medium", 
                                                      returnclass = "sf",
                                                      country = c("Malaysia", "Indonesia")))

## prepare land cover layer (year 2020) and crop for the region
if(!file.exists("../../data/supply_shed_data/worldcover_2020_cropped.tif")) {
  World_Cover_2020 <-
    rast("../../data/supply_shed_data/worldcover_2020_idn_mys_bias_oilpalm_cog.tif")
  World_Cover_2020_crop <-
    crop(World_Cover_2020, Indonesia_Malaysia_Map)
  writeRaster(World_Cover_2020_crop,
              '../../data/supply_shed_data/worldcover_2020_cropped.tif')
  rm(World_Cover_2020)
} else{
  World_Cover_2020_crop <-
    rast('../../data/supply_shed_data/worldcover_2020_cropped.tif')
}

## prepare land cover layer (year 2021) and crop for the region
if(!file.exists("../../data/supply_shed_data/worldcover_2021_cropped.tif")) {
  World_Cover_2021 <-
    rast("../../data/supply_shed_data/worldcover_2021_idn_mys_bias_oilpalm_cog.tif")
  World_Cover_2021_crop <-
    crop(World_Cover_2021, Indonesia_Malaysia_Map)
  writeRaster(World_Cover_2021_crop,
              '../../data/supply_shed_data/worldcover_2021_cropped.tif')
  rm(World_Cover_2021)
} else{
  World_Cover_2021_crop <-
    rast('../../data/supply_shed_data/worldcover_2021_cropped.tif')
}

## to examine the land cover categories: 
## see https://developers.google.com/earth-engine/datasets/catalog/ESA_WorldCover_v200#bands for the values of the bands and corresponding land use category; 11 was added to encoded oil palm plantations.

# Mill_buffer_25 <- rast("../data/supply_shed_data/UML_data/mill_radius_25km.tif")
# 
# UML <- read.csv('../data/supply_shed_data/UML_data/UML-Dec-2024.csv', sep=',', fileEncoding="latin1") %>% 
#  filter(Country %in% c('Malaysia', 'Indonesia'))
# 
# v <- vect(matrix(c(as.numeric(UML$Longitude[1]),
#             as.numeric(UML$Latitude[1])), ncol=2),
#           crs="+proj=longlat +datum=WGS84")
# v_buffer <- buffer(v, width=25000)
# 
# r_cropped <- crop(World_Cover_2021_crop, v_buffer)
# r_masked <- mask(r_cropped, v_buffer)
# plot(as.factor(r_masked))
```

```{r}
#| output: false 
#| echo: true 
#| label: loading and preparing the forest integrity layer (this is for 2019)
## this layer is from https://www.nature.com/articles/s41467-020-19493-3#Sec4
Forest_integrity <- rast('../../data/Forest_integrity/flii_Asia.tif')
Forest_integrity_crop <-
    crop(Forest_integrity, Indonesia_Malaysia_Map)
Forest_integrity_crop <- Forest_integrity_crop/1000
#plot(Forest_integrity_crop)
```

## 1.3. Load the EO data used as explanatory variables in the predictive models and other relevant EO data

```{r}
#| output: false 
#| echo: true 
#| label: loading and preparing the EO layers used in the model fitting

if(!file.exists('../../results/1_extraction_at_sites/site_level_extractions_Danylo_500.csv')){

## defining a cylindrical equal area projection centered in Indonesia Malaysia
cea_crs_custom <- crs('+proj=cea +lon_0=110 +lat_ts=0 +x_0=0 +y_0=0 +datum=WGS84 +ellps=WGS84 +units=m +no_defs')

## loading EO data data:

####### EO data (in geographic crs): Danylo et al. - oil palm plantation age
Tif_files_list <- list.files("P:/bec_data/100_rawdata/oil_palm_Danylo_2021/oil_palm_detection_year_Danylo_data/", pattern = "\\.tif$", full.names = TRUE)
vrt_Danylo <- terra::vrt(Tif_files_list)
rm(Tif_files_list)

# ####### EO data (in geographic crs): Descals et al. - oil palm planting year (possible alternative to Danylo et al.)
# Tif_files_list <- list.files("P:/bec_data/100_rawdata/oil_palm_Descals_2024/GlobalOilPalm_OP-YoP/", pattern = "\\.tif$", full.names = TRUE)
# vrt_Descals_yop <- terra::vrt(Tif_files_list)
# rm(Tif_files_list)

####### EO data: ESA CCI land cover maps / unprojected
## loading 24 annual LC maps from 1992 to 2015, from ESA CCI 
LC_maps <- terra::rast("../../data/ESACCI-LC-L4-LCCS-Map-300m-P1Y-1992_2015-v2.0.7.tif")
LC_legend <- read.csv('../../data/ESACCI-LC-legend.csv', sep=";")
names(LC_maps) <- as.character(1992:2015)

####### EO data: Intact forest landscape (baseline years 2000 and 2013)
IFL_2000 <- terra::vect("P:/bec_data/100_rawdata/IntactForestLandscapes/IFL_2000/ifl_2000.shp")
IFL_2013 <- terra::vect("P:/bec_data/100_rawdata/IntactForestLandscapes/IFL_2013/ifl_2013.shp")

raster_template <- crop(LC_maps[[1]], Indonesia_Malaysia_Map)

IFL_2000_crop_rast <- rasterize(crop(IFL_2000, Indonesia_Malaysia_Map), raster_template, field="AREA2020")
IFL_2000_crop_rast[IFL_2000_crop_rast > 0] <- 1

IFL_2013_crop_rast <- rasterize(crop(IFL_2013, Indonesia_Malaysia_Map), raster_template, field="AREA2020")
IFL_2013_crop_rast[IFL_2013_crop_rast > 0] <- 1

IFL_layers <- c(IFL_2000_crop_rast, IFL_2013_crop_rast)
names(IFL_layers) <- c('2000', '2013')
rm(Indonesia_Malaysia_Map, IFL_2000, IFL_2000_crop_rast, IFL_2013, IFL_2013_crop_rast)

####### EO data: Austin et al. maps - industrial oil palm plantations (versus smallholders) / https://pure.iiasa.ac.at/id/eprint/14829/
Austin_2017 <- list.files("P:/bec_data/100_rawdata/oil_palm_Austin_2017/LUPpaper2017_indo_oilpalm_map/IIASA_indo_oilpalm_map/", pattern = "\\.tif$", full.names = TRUE)
Austin_2017 <- lapply(Austin_2017, rast)

Austin_2017_stack <- c()
for(i in 1:length(Austin_2017)){
  rasti <- Austin_2017[[i]]
  ext(rasti) <- ext(raster_template)
  Austin_2017_stack <- c(Austin_2017_stack, rasti)
}
names(Austin_2017_stack) <- c('1995', '2000', '2005', '2010', '2015')

rm(raster_template, i, rasti, Austin_2017)

}
```

## 1.4. Load and prepare the PREDICTS data, subset for Indonesia and Malaysia

The PREDICTS data are loaded and pre-processed by correcting measurements for sampling effort and merging sites. We use the StatisticalModels package authored by Tim Newbold (UCL).

```{r}
#| output: false 
#| echo: true 

if(!file.exists('../../results/0_processed_data/PREDICTS_preprocessed.rds')){

  ## PREDICTS database (2016 v1.1) & 2022 additional data: bind together
  PREDICTS_2016_v1.1 <- readRDS('P:/bec_data/100_rawdata/PREDICTS/2016_V1.1/database.rds')
  Add_data <- readRDS('P:/bec_data/100_rawdata/PREDICTS/2022/database_2nd_release_12_11_2025.rds')
  PREDICTS <- rbind(PREDICTS_2016_v1.1, Add_data)
  rm(PREDICTS_2016_v1.1, Add_data)
  
  ## pre-process the PREDICTS database: correct for sampling effort and merge sites using functions from Tim Newbold
  PREDICTS <- predictsFunctions::CorrectSamplingEffort(diversity = PREDICTS)
  PREDICTS <- predictsFunctions::MergeSites(diversity = PREDICTS, silent = TRUE)
  saveRDS(PREDICTS, '../../results/0_processed_data/PREDICTS_preprocessed.rds')
  
} else{
  
  if(file.exists('../../results/0_processed_data/PREDICTS_preprocessed_Indonesia_Malaysia.rds')){
    
    PREDICTS_Indonesia_Malaysia <- readRDS('../../results/0_processed_data/PREDICTS_preprocessed_Indonesia_Malaysia.rds')
    
  } else{
    
    PREDICTS <- readRDS('../../results/0_processed_data/PREDICTS_preprocessed.rds')
    
    ## Subset the data for Indonesia and Malaysia, which constitute the pilot region. Add detailed habitat information: for 
    ## flag sites identified as oil palm plantations.
    PREDICTS_Indonesia_Malaysia <-  subset(PREDICTS,  Country %in% c('Indonesia', 'Malaysia'))
    PREDICTS_Indonesia_Malaysia$Country <- droplevels(PREDICTS_Indonesia_Malaysia$Country)
    
    PREDICTS_Indonesia_Malaysia$is_oil_palm_PREDICTS <- NA
    PREDICTS_Indonesia_Malaysia$is_oil_palm_PREDICTS[grepl(
      'oil palm stand|oil palm managed|Oil Palm Plantation|oil-palm plantations|Oil palm plantation|Palm oil plantation|Monoculture palm oil',
      PREDICTS_Indonesia_Malaysia$Habitat_as_described)] <- TRUE
    PREDICTS_Indonesia_Malaysia$is_oil_palm_PREDICTS[is.na(PREDICTS_Indonesia_Malaysia$is_oil_palm_PREDICTS)] <- FALSE
    
    # table(PREDICTS_Indonesia_Malaysia$is_oil_palm_PREDICTS)
    # OtherLU <- unique(droplevels(PREDICTS_Indonesia_Malaysia$Habitat_as_described)[is.na(PREDICTS_Indonesia_Malaysia$is_oil_palm_PREDICTS)])
    # OtherLU[grepl('oil', OtherLU)]
    
    saveRDS(PREDICTS_Indonesia_Malaysia, '../../results/0_processed_data/PREDICTS_preprocessed_Indonesia_Malaysia.rds')
    rm(PREDICTS)
  }
}
```

Derive site-level information for extracting EO-derived indicators around sites.

```{r}
#| output: false 
#| echo: true 
All_sites <-
  unique(PREDICTS_Indonesia_Malaysia[, c(
    'SS',
    'SSB',
    'SSBS',
    'Longitude',
    'Latitude',
    'Predominant_land_use',
    'Sample_start_earliest',
    'Sample_end_latest',
    'Sample_midpoint',
    "Max_linear_extent_metres",
    "Habitat_patch_area_square_metres",
    "Habitat_as_described",
    'is_oil_palm_PREDICTS'
  )])

rownames(All_sites) <- c(1:nrow(All_sites))
```

## 1.5. Extract EO-derived indices around sampled sites:

We use the data from Danylo et al. to extract plantation age around each site. For each site, clip the virtual mosaic to a specified buffer area around the site (here we use 500 metres), and extract the value of the EO layer. Where there are negative values, allocate NAs (sampling cannot have happened before planting therefore the site is not likely to be an oil palm plantation).

```{r}
#| output: false 
#| echo: true 

## the extraction takes about 2 hours (Here, extracting age from the Danylo et al. layer)

if(file.exists('../../results/1_extraction_at_sites/Danylo_age_500m.csv')) {
  Results_Danylo <-
    read.csv('../../results/1_extraction_at_sites/Danylo_age_500m.csv')
} else{
  ## Plantation age derived from the Danylo layer -- using a buffer of 500m
  Results_Danylo <-
    Loop_over_sites_Extract_age(
      Sites_data = All_sites,
      virtual_mosaic = vrt_Danylo,
      buffer_crop_area = 5000, ## clip tp a 5km buffer zone before projecting
      buffer_extract_data = 500, ## 500m buffer for extracting data
      baseline_year = 1980, ## for calculations using Danylo layer
      EO_Layer = 'Danylo'
    )
  
  write.csv(
    Results_Danylo,
    '../../results/1_extraction_at_sites/Danylo_age_500m.csv',
    row.names = FALSE
  )
  
}
```

Extract for additional EO-layers: land cover, intact forest landscape, industrial oil palm plantations.

```{r}
#| output: false 
#| echo: true 
if(file.exists('../../results/1_extraction_at_sites/Danylo_500m.csv')){
  Run_Danylo_500 <- read.csv('../../results/1_extraction_at_sites/Danylo_500m.csv')
} else {
## using a 500m buffer for extraction from EO layers => circle of 1km around site
Run_Danylo_500 <- Loop_over_sites_Extract_LC(Sites_data = All_sites, 
                                             ESA_CCI_layers = LC_maps, 
                                             EO_age_layer = 'Danylo',
                                             EO_age_detection_layer = vrt_Danylo,
                                             IFL_layer = IFL_layers, 
                                             Austin_OP_layer = Austin_2017_stack,
                                             buffer_crop_area = 5000,
                                             buffer_extract_data = 500, 
                                             Save_Plot = FALSE)

write.csv(Run_Danylo_500, '../../results/1_extraction_at_sites/Danylo_500m.csv', row.names=FALSE)
}


## consolidate the extracted data together, and identify oil palm plantations 
Results_Danylo <- Results_Danylo %>% 
  dplyr::select(SSBS, Danylo_linear_extent_age_proj_median, Danylo_age_cea) %>% 
  setNames(., c('SSBS', 'Age_LE_median', 'Age_exact_location'))
Run_Danylo_500 <- left_join(Run_Danylo_500, Results_Danylo)
```

Identify sites which we assume are oil-palm plantations. We assume plantations when the age extracted along the site linear extent is \>0. Set factors levels for land uses and save results.

```{r}
#| output: false 
#| echo: true

## identif oil palm plantations
Run_Danylo_500$is_oil_palm_Danylo <- NA
Run_Danylo_500$is_oil_palm_Danylo[Run_Danylo_500$Age_LE_median>0] <- TRUE

## set factor levels for land use and save results
Run_Danylo_500$Predominant_land_use <- factor(
  Run_Danylo_500$Predominant_land_use,
  levels = c(
    "Primary vegetation",
    "Mature secondary vegetation",
    "Intermediate secondary vegetation",
    "Young secondary vegetation",
    "Secondary vegetation (indeterminate age)",
    "Plantation forest",
    "Pasture",
    "Cropland",
    "Urban",
    "Cannot decide"
  )
)

write.csv(Run_Danylo_500, '../../results/1_extraction_at_sites/site_level_extractions_Danylo_500.csv')
```

## 1.6. Derive the site-level biodiversity metrics

We first calculate site-level total abundance, and we rescale it (within studies). We also derive species richness.

```{r}
#| output: false 
#| echo: true
#| label: site-level metrics (rescaled abundance, species richness)

## calculating site-level abundance (we get total abundance and rescaled total abundance)
site_total_abundance <- PREDICTS_Indonesia_Malaysia %>% 
  filter(Diversity_metric_type=='Abundance') %>% 
  group_by(SS, SSBS) %>% 
  summarise(total_abundance=sum(Measurement)) %>% 
  ungroup() %>% 
  group_by(SS) %>% 
  mutate(max_abundance = max(total_abundance)) %>% 
  ungroup() %>% 
  mutate(total_abundance_rescaled = total_abundance/max_abundance) %>%  
  ungroup() %>% 
  dplyr::select(-max_abundance)

## species richness
## calculating site-level species richness 
SR_1 <- PREDICTS_Indonesia_Malaysia %>% 
  filter(Diversity_metric_type=='Species richness') %>% 
  group_by(SS, SSBS) %>% 
  summarise(SR=sum(Measurement))
SR_2 <- PREDICTS_Indonesia_Malaysia %>% 
  filter(Diversity_metric_type %in% c('Abundance','Occurrence')) %>% 
  mutate(PresenceAbsence=ifelse(Measurement>0, 1, 0)) %>% 
  group_by(SS, SSBS) %>% 
  summarise(SR=sum(PresenceAbsence))
site_species_richness <- rbind(SR_1, SR_2)
rm(SR_1, SR_2)

## join the site-level biodiversity metrics with the site-level data
All_sites <- left_join(Run_Danylo_500, site_total_abundance)
All_sites <- left_join(All_sites, site_species_richness)

## save these site-level data
write.csv(All_sites, '../../results/2_biodiversity_metrics_site_level/site_level_metrics_Danylo_500.csv', row.names=FALSE)
```

Now, we calculate dissimilarity metrics, which relies on pairwise comparisons beween sites within studies. Within each study, sites are matched with all reference primary vegetation sites. We first need to derive pair of sites. In each pair, disturbed land-use types are compared with (minimally-disturbed) primary vegetation, where pairs of sites are available within the same study. We consider all primary vegetation sites here (not just minimally disturbed) in this version. First we create relevant pairs of sites.

```{r}
#| output: false 
#| echo: true
#| label: create site pairs

## add whether a site was found to be an oil palm plantation site in Danylo data
PREDICTS_Indonesia_Malaysia <- left_join(PREDICTS_Indonesia_Malaysia, 
                                         All_sites %>% 
                                           dplyr::select(SS, SSBS, is_oil_palm_Danylo))

## retrieving relevant study ID for creating pairs of sites: 
## studies need to have at least one sampled primary vegetation sites
## studies need to have at least one sampled species
## putting together pairs of sites with measured abundances, for computation of Bray-Curtis dissimilarity index
## the reference sites are primary vegetation sites. Usually, we would use minimally-used PV as the reference, but we don't have any indication regarding intensity of the landscape from the land cover layers we will use for projections, so we use all categories instead.
Abundance_study_IDs <- 
  PREDICTS_Indonesia_Malaysia %>%
  filter(Diversity_metric_type=="Abundance") %>%
  dplyr::select(SS, SSB, SSBS, Predominant_land_use, Taxon_name_entered, Best_guess_binomial, Sampling_effort, is_oil_palm_Danylo) %>%
  # group by study
  group_by(SS) %>%
  # check whether there are primary vegetation sites within the studies (need at least one pv site)
  dplyr::mutate(n_pv_sites = sum(Predominant_land_use == "Primary vegetation")) %>%
  # also add number of taxa sampled in each study
  dplyr::mutate(n_taxa = dplyr::n_distinct(Taxon_name_entered)) %>% 
  # # number of unique sampling efforts within studies (need unique sampling effort within studies)
  dplyr::mutate(n_sampling_effort = dplyr::n_distinct(Sampling_effort)) %>% 
  ungroup() %>% 
  #  keep only studies with at least one Primary vegetation site
  dplyr::filter(n_pv_sites > 0) %>% 
  #  keep only studies with more than one taxa sampled
  dplyr::filter(n_taxa > 1) %>% 
  droplevels()

## retrieve the studies ID that are retained  
Abundance_study_IDs <- Abundance_study_IDs %>% 
  dplyr::distinct(SS)  %>% 
  dplyr::pull() %>% 
  as.character()

## about 76% of the studies are retained at this stage.
length(Abundance_study_IDs)/length(unique(PREDICTS_Indonesia_Malaysia$SS[PREDICTS_Indonesia_Malaysia$Diversity_metric_type=='Abundance']))*100

## Create relevant pairs for abundance
Abundance_pairs_data <- lapply(X=as.list(Abundance_study_IDs), 
                               FUN = CreatePairs,
                               input_data=PREDICTS_Indonesia_Malaysia) %>% 
  data.table::rbindlist() %>% 
  as.data.frame()

## filter out sites that have more than one sampling effort
Sites_sampling_efforts <- unique(PREDICTS_Indonesia_Malaysia[, c('SS', 'SSBS', 'Sampling_effort')])
Sites_sampling_efforts_n <- Sites_sampling_efforts |> 
  group_by(SSBS) |> 
  summarise(N_samling_effort = dplyr::n_distinct(Sampling_effort))
Sites_to_exclude <- Sites_sampling_efforts_n$SSBS[Sites_sampling_efforts_n$N_samling_effort>1]

Abundance_pairs_data_filtered <- Abundance_pairs_data |> 
  filter(!(SSBS_1 %in% Sites_to_exclude | SSBS_2 %in% Sites_to_exclude))
rm(Sites_sampling_efforts_n, Sites_to_exclude)

## add sampling effort of each pair of sites
Abundance_pairs_data_filtered <- left_join(Abundance_pairs_data_filtered, 
                                           Sites_sampling_efforts |> 
                                             mutate(SSBS_1=SSBS) |> 
                                             dplyr::select(SSBS_1, Sampling_effort))
colnames(Abundance_pairs_data_filtered)[colnames(Abundance_pairs_data_filtered)=='Sampling_effort'] <- 'Sampling_effort_SSBS_1'

Abundance_pairs_data_filtered <- left_join(Abundance_pairs_data_filtered, 
                                           Sites_sampling_efforts |> 
                                             mutate(SSBS_2=SSBS) |> 
                                             dplyr::select(SSBS_2, Sampling_effort))
colnames(Abundance_pairs_data_filtered)[colnames(Abundance_pairs_data_filtered)=='Sampling_effort'] <- 'Sampling_effort_SSBS_2'

Abundance_pairs_data_filtered <- Abundance_pairs_data_filtered |> 
  filter(Sampling_effort_SSBS_1==Sampling_effort_SSBS_2)

## add sites identified as oil palm plantations, and create contrasts
Abundance_pairs_data_filtered <- left_join(Abundance_pairs_data_filtered, 
                                           All_sites |> 
                                             mutate(SSBS_2=SSBS) |> 
                                             dplyr::select(SSBS_2, is_oil_palm_Danylo))

Abundance_pairs_data_filtered$contrast <- ifelse(is.na(Abundance_pairs_data_filtered$is_oil_palm_Danylo), 
                                                 paste(Abundance_pairs_data_filtered$lu_1, Abundance_pairs_data_filtered$lu_2, sep='-'), 
                                                 paste(Abundance_pairs_data_filtered$lu_1, 'Oil palm plantation', sep='-'))

Abundance_pairs_data_filtered$contrast <- factor(Abundance_pairs_data_filtered$contrast, 
                                                 levels=c("Primary vegetation-Primary vegetation", 
                                                          "Primary vegetation-Mature secondary vegetation",
                                                          "Primary vegetation-Intermediate secondary vegetation", 
                                                          "Primary vegetation-Young secondary vegetation", 
                                                          "Primary vegetation-Secondary vegetation (indeterminate age)",
                                                          "Primary vegetation-Plantation forest", 
                                                          "Primary vegetation-Cropland",
                                                          "Primary vegetation-Oil palm plantation"))

#table(Abundance_pairs_data_filtered$contrast)
```

Now, we calculate the dissimilarity metric (Bray-Curtis):

```{r}
#| output: false 
#| echo: true
#| label: calculate Bray Curtis dissimilary 
## it takes ~1.5h
if(!file.exists('../../results/2_biodiversity_metrics_site_level/dissimilarity_metrics_Danylo_500.csv')){
Bray_Cutis_dissimilarity_vegan <-
  mapply(
    FUN = Calculate_Bray_Curtis_dissimilarity,
    site_1 = as.list(Abundance_pairs_data_filtered$SSBS_1),
    site_2 = as.list(Abundance_pairs_data_filtered$SSBS_2),
    MoreArgs = list(input_data =
                      PREDICTS_Indonesia_Malaysia),
    SIMPLIFY = FALSE
  )

Bray_Cutis_dissimilarity_vegan <- data.table::rbindlist(Bray_Cutis_dissimilarity_vegan)
Bray_Cutis_dissimilarity_vegan <- left_join(Abundance_pairs_data_filtered, Bray_Cutis_dissimilarity_vegan)

## save these dissimilarity estimates
write.csv(Bray_Cutis_dissimilarity_vegan, '../../results/2_biodiversity_metrics_site_level/dissimilarity_metrics_Danylo_500.csv', row.names=FALSE)
} else{
  Bray_Cutis_dissimilarity_vegan <- read.csv('../../results/2_biodiversity_metrics_site_level/dissimilarity_metrics_Danylo_500.csv')
}


## transform the Bray-Curtis similarity with a logit 
Bray_Cutis_dissimilarity_vegan$logit_bray_curtis_similarity_vegan <- car::logit(1-Bray_Cutis_dissimilarity_vegan$bray_curtis_distance_vegan, adjust = 0.001, percents = FALSE)
## log10 geographic distance
Bray_Cutis_dissimilarity_vegan$log10_geodist <- log10(Bray_Cutis_dissimilarity_vegan$site_distance_meters+1)
```

# 2. ANALYSES AND PREDICTIVE MODEL FITTING

We have now obtained a dataset containing biodiversity records and metrics for Malaysia and Indonesia, specifically sampled in primary and secondary vegetation, oil-palm plantations, and other land-use types. Now we fit the predictive statistical models.

## 2.1. Total-abundance models

We first apply some transformations to the total rescaled abundance data to improve normality. We consider a square-root transformation.

```{r}
#| output: false 
#| echo: true 
All_sites$sqrt_total_abundance_rescaled <- sqrt(All_sites$total_abundance_rescaled)
#hist(All_sites$sqrt_total_abundance_rescaled )

## we set the % share of CII LU to 0 where it is NA (the LU was not detected)
All_sites$LU_CCI_50[is.na(All_sites$LU_CCI_50)] <- 0
```

Now we fit a mixed-effects model to investigate whether oil palm age affects abundance.

```{r}
## effect of age on abundance or dissimilarity 
#| output: false 
#| echo: true 
#plot(All_sites$sqrt_total_abundance_rescaled[All_sites$is_oil_palm_Danylo]~ All_sites$Median_age[All_sites$is_oil_palm_Danylo], pch=19) 

## no effect of age on rescaled abundance in any of the following models considered 
summary(lme4::lmer(formula = sqrt_total_abundance_rescaled ~ Median_age + (1|SS / SSB), data = All_sites[All_sites$is_oil_palm_Danylo, ]))
summary(lme4::lmer(formula = sqrt_total_abundance_rescaled ~ MAD_age + (1 |SS / SSB), data = All_sites[All_sites$is_oil_palm_Danylo, ]))
summary(lme4::lmer(formula = sqrt_total_abundance_rescaled ~ Age_LE_median + (1 |SS / SSB), data = All_sites[All_sites$is_oil_palm_Danylo, ]))
summary(lme4::lmer(formula = sqrt_total_abundance_rescaled ~ Age_exact_location + (1 |SS / SSB), data = All_sites[All_sites$is_oil_palm_Danylo, ])) 

# ## effect on dissimilarity
# OP_diss <- left_join(Bray_Cutis_dissimilarity_vegan, 
#                      All_sites[, c('SSBS', 
#                                    'Median_age', 
#                                    'MAD_age', 
#                                    'Age_LE_median',
#                                    'Age_exact_location')] %>% 
#                        rename(SSBS_2=SSBS)) %>% 
#   filter(is_oil_palm_Danylo)
# 
# plot(OP_diss$logit_bray_curtis_similarity_vegan~OP_diss$MAD_age)
# 
# ## fit similarity models
# summary(
# lme4::lmer(
#   logit_bray_curtis_similarity_vegan ~  Median_age + log10_geodist + (1|SS), 
#   data = OP_diss
#   ))
# summary(
# lme4::lmer(
#   logit_bray_curtis_similarity_vegan ~  Age_exact_location + log10_geodist + (1|SS), 
#   data = OP_diss
#   ))
# summary(
# lme4::lmer(
#   logit_bray_curtis_similarity_vegan ~  MAD_age + log10_geodist + (1|SS), 
#   data = OP_diss
#   ))
# summary(
# lme4::lmer(
#   logit_bray_curtis_similarity_vegan ~  Age_LE_median + log10_geodist + (1|SS), 
#   data = OP_diss
#   ))
```

Now we fit mixed-effects models investigating how land-use type affects abundance. Here, we consider all land-use types together.

```{r}
#| output: false 
#| echo: true
# Chunk 16
#| output: false 
#| echo: true

## checking the land-use categories
All_sites$LandUse <- as.character(All_sites$Predominant_land_use)
All_sites$LandUse[All_sites$is_oil_palm_Danylo] <- 'Oil palm plantation'
All_sites$LandUse[grepl('econdary vegetation',All_sites$Predominant_land_use)] <- 'Secondary vegetation'
All_sites <- All_sites[All_sites$LandUse!='Urban',]
All_sites <- All_sites[All_sites$LandUse!='Cannot decide',]

All_sites$LandUse <- factor(
  All_sites$LandUse,
  levels = c(
    "Primary vegetation",
    "Secondary vegetation",
    "Plantation forest",
    'Oil palm plantation',
    "Pasture",
    "Cropland",
    "Urban"
  )
)
All_sites$LandUse <- droplevels(All_sites$LandUse)
table(All_sites$LandUse)

## We also create a  new land use category that distinguishes different secondary vegetation age
All_sites$LandUse_SVcat <- as.character(All_sites$LandUse)
All_sites$LandUse_SVcat[All_sites$LandUse=='Secondary vegetation'] <- as.character(All_sites$Predominant_land_use[All_sites$LandUse=='Secondary vegetation'])
All_sites$LandUse_SVcat <- factor(
  All_sites$LandUse_SVcat,
  levels = c(
    "Primary vegetation",
    "Mature secondary vegetation",
    "Intermediate secondary vegetation",
    "Young secondary vegetation",
    "Secondary vegetation (indeterminate age)",
    "Plantation forest",
    'Oil palm plantation',
    "Cropland")
)

table(All_sites$LandUse_SVcat)


########################################## Models without secondary vegetation age
## first model with total rescaled abundance
abundance_model <- lme4::lmer(formula = sqrt_total_abundance_rescaled ~ LandUse  + (1|SS/SSB), data=All_sites)
summary(abundance_model)

## second model with total rescaled abundance
abundance_model_treecover <- lme4::lmer(formula = sqrt_total_abundance_rescaled ~ LandUse  + LU_CCI_50 + (1|SS/SSB), data=All_sites)
summary(abundance_model_treecover)

## which model fits better? the model with tree cover
anova(abundance_model, abundance_model_treecover)

## does adding oil palm plantation age make the model fit better? No
abundance_model_treecover_2 <- lme4::lmer(formula = sqrt_total_abundance_rescaled ~ LandUse  + LU_CCI_50 + (1|SS/SSB), data=All_sites[!is.na(All_sites$Median_age),])

abundance_model_treecover_age <- lme4::lmer(formula = sqrt_total_abundance_rescaled ~ LandUse  + LU_CCI_50 + Median_age + (1|SS/SSB), data=All_sites[!is.na(All_sites$Median_age),])

anova(abundance_model_treecover_2, abundance_model_treecover_age)
rm(abundance_model_treecover_2, abundance_model_treecover_age)

## does adding land-use intensity make the model fit better?
All_sites_LUI <- left_join(All_sites, unique(PREDICTS_Indonesia_Malaysia[, c('SSBS', 'Use_intensity')])) 
All_sites_LUI <- subset(All_sites_LUI, !is.na(Use_intensity))
All_sites_LUI <- subset(All_sites_LUI, Use_intensity!='Cannot decide')
All_sites_LUI$Use_intensity <- droplevels(All_sites_LUI$Use_intensity)

abundance_model_treecover_2 <- lme4::lmer(formula = sqrt_total_abundance_rescaled ~ LandUse  + LU_CCI_50 + (1|SS/SSB), data=All_sites_LUI)

abundance_model_treecover_intensity <- lme4::lmer(formula = sqrt_total_abundance_rescaled ~ LandUse  + LU_CCI_50 + Use_intensity + (1|SS/SSB), data=All_sites_LUI)

anova(abundance_model_treecover_2, abundance_model_treecover_intensity)
rm(abundance_model_treecover_2)

## including land-use intensity improves the model. 

########################################## Models with secondary vegetation age
## first model with total rescaled abundance
abundance_model_SV <- lme4::lmer(formula = sqrt_total_abundance_rescaled ~ LandUse  + (1|SS/SSB), data=All_sites %>%
      filter(LandUse_SVcat != 'Secondary vegetation (indeterminate age)') %>%
      mutate(LandUse_SVcat=droplevels(LandUse_SVcat))%>% 
      mutate(LandUse=LandUse_SVcat))

## second model with total rescaled abundance
# abundance_model_SV_treecover <-
#   lme4::lmer(
#     formula = sqrt_total_abundance_rescaled ~ LandUse_SVcat  + LU_CCI_50 + (1 |
#                                                                               SS / SSB),
#     data = All_sites %>%
#       filter(LandUse_SVcat != 'Secondary vegetation (indeterminate age)') %>%
#       mutate(LandUse_SVcat=droplevels(LandUse_SVcat))
#   )

abundance_model_SV_treecover <-
  lme4::lmer(
    formula = sqrt_total_abundance_rescaled ~ LandUse  + LU_CCI_50 + (1 |
                                                                              SS / SSB),
    data = All_sites %>%
      filter(LandUse_SVcat != 'Secondary vegetation (indeterminate age)') %>%
      mutate(LandUse_SVcat=droplevels(LandUse_SVcat)) %>% 
      mutate(LandUse=LandUse_SVcat)
  )

## which model fits better? the model with tree cover fits better
anova(abundance_model_SV, abundance_model_SV_treecover)

## does adding land-use intensity make the model fit better?
abundance_model_SV_treecover_2 <-
  lme4::lmer(
    formula = sqrt_total_abundance_rescaled ~ LandUse + LU_CCI_50 + (1 |
                                                                             SS / SSB),
    data = All_sites_LUI %>%
       filter(LandUse_SVcat != 'Secondary vegetation (indeterminate age)') %>%
      mutate(LandUse_SVcat=droplevels(LandUse_SVcat)) %>% 
      mutate(LandUse=LandUse_SVcat)
  )

abundance_model_SV_treecover_intensity <-
  lme4::lmer(
    formula = sqrt_total_abundance_rescaled ~ LandUse  + LU_CCI_50 + Use_intensity + (1 |
                                                                                              SS / SSB),
    data = All_sites_LUI %>%
      filter(LandUse_SVcat != 'Secondary vegetation (indeterminate age)') %>%
      mutate(LandUse_SVcat=droplevels(LandUse_SVcat)) %>% 
      mutate(LandUse=LandUse_SVcat)
  )

anova(abundance_model_SV_treecover_2, abundance_model_SV_treecover_intensity)
rm(abundance_model_SV_treecover_2)
## no strong evidence that including land use intensity impoves the model when considering secondary vegetation stages
```

Drawing predictions for the best model and comparing effect sizes with and without considering land use intensity.

```{r}
## drawing predictions
#| output: false 
#| echo: true
## predictions for the model on rescaled abundance, without land use intensity
## model that does not consider different stages for secondary vegetation
Newdata <- expand.grid(LandUse=levels(All_sites$LandUse), 
                       LU_CCI_50=50)

abundance_predictions <- Predict_effects_impr(
  new_data = Newdata,
  Model = abundance_model_treecover,
  rescale = TRUE,
  LU_n = length(levels(All_sites$LandUse)),
  backtransform = 'sqrt'
)

abundance_predictions$Median <- abundance_predictions$Median - 100
abundance_predictions$Lower <- abundance_predictions$Lower - 100
abundance_predictions$Upper <- abundance_predictions$Upper - 100

rm(Newdata)

## predictions for the model on rescaled abundance, with land use intensity
## model that does not consider different stages for secondary vegetation
Newdata <- expand.grid(LandUse=levels(All_sites_LUI$LandUse), 
                       Use_intensity=levels(All_sites_LUI$Use_intensity),
                       LU_CCI_50=50)

abundance_predictions_LUI <- Predict_effects_impr(
  new_data = Newdata,
  Model = abundance_model_treecover_intensity,
  rescale = TRUE,
  LU_n = length(levels(All_sites_LUI$LandUse))  *
    length(levels(All_sites_LUI$Use_intensity)),
  backtransform = 'sqrt'
)

abundance_predictions_LUI$Median <- abundance_predictions_LUI$Median - 100
abundance_predictions_LUI$Lower <- abundance_predictions_LUI$Lower - 100
abundance_predictions_LUI$Upper <- abundance_predictions_LUI$Upper - 100

abundance_predictions_all <- rbind(abundance_predictions %>% 
                                     mutate(Use_intensity='No intensity'), 
                                   abundance_predictions_LUI)

abundance_predictions_all$Use_intensity <-
  factor(
    abundance_predictions_all$Use_intensity,
    levels = c('No intensity',
               'Minimal use',
               'Light use',
               'Intense use'),
    labels = c('No intensity in model',
               'Minimal use',
               'Light use',
               'Intense use')
  )

## predictions for the model on rescaled abundance, without land use intensity
## model that considers different stages for secondary vegetation
Newdata <-
  expand.grid(
    LandUse = levels(droplevels(All_sites$LandUse_SVcat[All_sites$LandUse_SVcat !=
                                                          'Secondary vegetation (indeterminate age)'])),
    LU_CCI_50 = 50
  )

abundance_predictions_SV <- Predict_effects_impr(new_data = Newdata,
                                         Model = abundance_model_SV_treecover, 
                                         rescale = TRUE,
                                         LU_n = length(levels(droplevels(All_sites$LandUse_SVcat[All_sites$LandUse_SVcat!='Secondary vegetation (indeterminate age)']))),
                                         backtransform = 'sqrt')

abundance_predictions_SV$Median <- abundance_predictions_SV$Median - 100
abundance_predictions_SV$Lower <- abundance_predictions_SV$Lower - 100
abundance_predictions_SV$Upper <- abundance_predictions_SV$Upper - 100

rm(Newdata)

## predictions for the model on rescaled abundance, with land use intensity
## model that considers different stages for secondary vegetation
Newdata <- expand.grid(LandUse=levels(droplevels(All_sites$LandUse_SVcat[All_sites$LandUse_SVcat!='Secondary vegetation (indeterminate age)'])), 
                       Use_intensity = levels(All_sites_LUI$Use_intensity),
                       LU_CCI_50=50)

abundance_predictions_SV_LUI <-
  Predict_effects_impr(
    new_data = Newdata,
    Model = abundance_model_SV_treecover_intensity,
    rescale = TRUE,
    LU_n = length(levels(droplevels(
      All_sites$LandUse_SVcat[All_sites$LandUse_SVcat != 'Secondary vegetation (indeterminate age)']
    ))) * length(levels(All_sites_LUI$Use_intensity)),
    backtransform = 'sqrt'
  )

abundance_predictions_SV_LUI$Median <- abundance_predictions_SV_LUI$Median - 100
abundance_predictions_SV_LUI$Lower <- abundance_predictions_SV_LUI$Lower - 100
abundance_predictions_SV_LUI$Upper <- abundance_predictions_SV_LUI$Upper - 100

rm(Newdata)

abundance_predictions_all_SV <- rbind(abundance_predictions_SV %>% 
                                     mutate(Use_intensity='No intensity'), 
                                   abundance_predictions_SV_LUI)

abundance_predictions_all_SV$Use_intensity <- factor(abundance_predictions_all_SV$Use_intensity, 
                                                  levels=c('No intensity', 
                                                           'Minimal use', 
                                                           'Light use', 
                                                           'Intense use'), 
                                                  labels=c('No intensity in model', 
                                                           'Minimal use', 
                                                           'Light use', 
                                                           'Intense use'))

## plotting effects: models that do not consider different stages of secondary vegetation
ggplot(abundance_predictions_all,
       aes(LandUse , Median, ymin = Lower, ymax = Upper, group=Use_intensity, col=Use_intensity)) +
  xlab("") +
  theme_classic()+
  geom_hline(yintercept = 0, col="black", linetype="dashed") +
  geom_errorbar(width=.2, size=0.5, stat="identity",position=position_dodge(width = 0.6)) +
  geom_point(size=2, position=position_dodge(width = 0.6)) +
  #GGPoptions +
  ylab("Abundance (% difference from primary)") + 
  theme(panel.spacing.y =  unit(0, "lines")) +
  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 12)) +
  scale_discrete_manual("colour",
                        values = c('black', 'red4', 'orange', 'darkgreen'),
                        breaks=c('No intensity in model', 'Intense use', 'Light use', 'Minimal use')) +
  ggtitle("Effects of models on total abundance (with and without land-use intensity)")

## plotting effects: models that consider different stages of secondary vegetation
ggplot(abundance_predictions_all_SV,
       aes(LandUse , Median, ymin = Lower, ymax = Upper, group=Use_intensity, col=Use_intensity)) +
  xlab("") +
  theme_classic()+
  geom_hline(yintercept = 0, col="black", linetype="dashed") +
  geom_errorbar(width=.2, size=0.5, stat="identity",position=position_dodge(width = 0.6)) +
  geom_point(size=2, position=position_dodge(width = 0.6)) +
  #GGPoptions +
  ylab("Abundance (% difference from primary)") + 
  theme(panel.spacing.y =  unit(0, "lines")) +
  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 12)) +
  scale_discrete_manual("colour",
                        values = c('black', 'red4', 'orange', 'darkgreen'),
                        breaks=c('No intensity in model', 'Intense use', 'Light use', 'Minimal use')) +
  ggtitle("Effects of models on total abundance (with and without land-use intensity)")
```

In summary:

-   When secondary vegetation is grouped into a single category (not considering different ages/stages), including land-use intensity improves the model. By using the model that does not consider land-use intensity, we likely under-estimate average biodiversity declines in some of the most disturbed land uses, as shown in plot above.

-   However, when considering secondary vegetation stages (mature/intermediate/young), there isn't strong evidence that including land-use intensity improves the model. We still likely under-estimate average biodiversity declines in the most disturbed (intensely-used) land uses, as shown in plot above.

-   Given that, for predictions, we don't have layers providing intensity for the different land uses, we choose the model that distinguishes different secondary vegetation types, without considering land use intensity. Indeed, we will assume that we can approximate secondary vegetation stage (for predictions) from the EO-derived forest integrity layer. We however don't have data at this stage on land-use intensity that we could use for prediction.

-   The abundance model we choose is therefore: 'abundance_model_SV_treecover'.

## 2.2. Similarity models

Since we already established that including land-use intensity in the model will be limiting (for deriving predictions), we don't consider the effects of land-use intensity on similarity. We likely underestimate the average changes in similarity in the most disturbed, intensely-used land use types.

```{r}
#| output: false 
#| echo: true
#| cache: false

Bray_Cutis_dissimilarity_vegan$contrast %>% unique()
# Bray_Cutis_dissimilarity_vegan$contrast[Bray_Cutis_dissimilarity_vegan$contrast %in% c("Primary vegetation-Mature secondary vegetation",
#                                                                                        "Primary vegetation-Intermediate secondary vegetation",
#     "Primary vegetation-Young secondary vegetation",
#     "Primary vegetation-Secondary vegetation (indeterminate age)")] <- "Primary vegetation-Secondary vegetation"

Bray_Cutis_dissimilarity_vegan <- subset(Bray_Cutis_dissimilarity_vegan, lu_2!='Secondary vegetation (indeterminate age)')

Bray_Cutis_dissimilarity_vegan$contrast <- factor(
  Bray_Cutis_dissimilarity_vegan$contrast,
    levels = c(
    "Primary vegetation-Primary vegetation",
    #"Primary vegetation-Secondary vegetation",
    "Primary vegetation-Mature secondary vegetation",
    "Primary vegetation-Intermediate secondary vegetation",
    "Primary vegetation-Young secondary vegetation",
    #"Primary vegetation-Secondary vegetation (indeterminate age)",
    "Primary vegetation-Plantation forest",
    "Primary vegetation-Oil palm plantation",
    "Primary vegetation-Cropland"
  )
)
table(Bray_Cutis_dissimilarity_vegan$contrast)

## fit similarity model
similarity_model <- lme4::lmer(
  logit_bray_curtis_similarity_vegan ~ contrast + log10_geodist + (1|SS), 
  data = Bray_Cutis_dissimilarity_vegan
  )

## fit similarity model with tree cover
Bray_Cutis_dissimilarity_vegan <- left_join(Bray_Cutis_dissimilarity_vegan, 
                      All_sites[, c('SSBS', 
                                    'LU_CCI_50')] %>% 
                        rename(SSBS_2=SSBS))

similarity_model_tree_cover <- lme4::lmer(
  logit_bray_curtis_similarity_vegan ~ contrast + LU_CCI_50 + log10_geodist + (1|SS), 
  data = Bray_Cutis_dissimilarity_vegan
  )

## which model fits better? the model with tree cover
anova(similarity_model, similarity_model_tree_cover)

```

Drawing predictions for the similarity model.

```{r}
#| output: false 
#| echo: true
Newdata <- expand.grid(contrast=levels(Bray_Cutis_dissimilarity_vegan$contrast),
                       log10_geodist=0, 
                       LU_CCI_50=50)

similarity_predictions <- Predict_effects_impr(new_data = Newdata,
                                        Model = similarity_model_tree_cover,
                                        rescale = TRUE, 
                                        LU_n = length(levels(Bray_Cutis_dissimilarity_vegan$contrast)), 
                                        backtransform = 'logit',
                                        A_logit = 0.001)

similarity_predictions$Median <- similarity_predictions$Median - 100
similarity_predictions$Lower <- similarity_predictions$Lower - 100
similarity_predictions$Upper <- similarity_predictions$Upper - 100
rm(Newdata)
```

# 3. RESULTS

## 3.1. Fitted data

The abundance model was fitted on 1198 sites sampled within 38 studies.

```{r}
#| output: true 
#| echo: true
## sample sizes
Fitted_abundance <- All_sites %>% 
               filter(SS %in%  abundance_model_SV_treecover@frame$SS)
nrow(Fitted_abundance)
nrow(abundance_model_SV_treecover@frame)

Fitted_abundance$SS %>% 
  unique() %>% 
  length()

Fitted_abundance$SSBS %>% 
  unique() %>% 
  length()
```

```{r}
#| output: true 
#| echo: true
## sample sizes
Fitted_similarity <- Bray_Cutis_dissimilarity_vegan %>% 
               filter(SS %in%  similarity_model_tree_cover@frame$SS)
nrow(Fitted_similarity)
nrow(similarity_model_tree_cover@frame)

Fitted_similarity$SS %>% 
  unique() %>% 
  length()

Fitted_similarity$contrast %>% 
  table() 
```

```{r}
#| output: true 
#| echo: true
## map sites
Indonesia_Malaysia_Map <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf", country = c("Malaysia", "Indonesia"))

ggplot2::ggplot(data = Indonesia_Malaysia_Map) + xlab("") + ylab("") +
  geom_sf(fill="lightgrey", colour="lightgrey") +
  geom_point(data=Fitted_abundance %>% 
               filter(Predominant_land_use!='Cannot decide'),
             aes(y=Latitude, x=Longitude, col=LandUse),
             pch=19,alpha=0.6) +
  theme_minimal() +
  theme(axis.text.x=element_blank(), 
        axis.text.y=element_blank(),
        text = element_text(size=15, family="serif"), 
        legend.text=element_text(size=16), 
        legend.position = 'bottom') +
  ggtitle('Biodiversity sampling sites') + 
  scale_colour_discrete(name = "Land use") +
guides(color = guide_legend(nrow = 2)) 


## sampling dates
ggplot(Fitted_abundance, 
       aes(as.Date(Sample_start_earliest))) +
  geom_histogram(binwidth = 130, col='darkgrey', fill='lightblue') + theme_bw() +
  xlab('Start date of biodiversity sampling') + ylab('Number of sampled sites') +
  scale_x_date(date_breaks = "2 year", date_labels =  "%Y") +
  theme(axis.text.x=element_text(angle=60, hjust=1, size=15, face='bold'), 
        axis.text.y=element_text(size=15),
        text = element_text(size=15, family="serif", face='bold'), 
        legend.text=element_text(size=15))


```

## 3.2. Abundance and similarity models: effects of land use

Oil-palm plantations exhibited average declines in total abundance of about 15% compared to primary vegetation. Similarity in oil-palm plantations exhibited average declines of about 43% when compared to primary vegetation.

```{r}
#| output: true 
#| echo: true

# ## plotting effects: models that do not consider different stages of secondary vegetation
# ggplot(abundance_predictions_all,
#        aes(LandUse , Median, ymin = Lower, ymax = Upper, group=Use_intensity, col=Use_intensity)) +
#   xlab("") +
#   theme_classic()+
#   geom_hline(yintercept = 0, col="black", linetype="dashed") +
#   geom_errorbar(width=.2, size=0.5, stat="identity",position=position_dodge(width = 0.6)) +
#   geom_point(size=2, position=position_dodge(width = 0.6)) +
#   #GGPoptions +
#   ylab("Abundance (% difference from primary)") + 
#   theme(panel.spacing.y =  unit(0, "lines")) +
#   theme(axis.text.x = element_text(angle = 60, hjust=1, size = 12)) +
#   scale_discrete_manual("colour",
#                         values = c('black', 'red4', 'orange', 'darkgreen'),
#                         breaks=c('No intensity in model', 'Intense use', 'Light use', 'Minimal use')) +
#   ggtitle("Effects of models on total abundance (with and without land-use intensity)")
# 
# ## plotting effects: models that consider different stages of secondary vegetation
# ggplot(abundance_predictions_all_SV,
#        aes(LandUse , Median, ymin = Lower, ymax = Upper, group=Use_intensity, col=Use_intensity)) +
#   xlab("") +
#   theme_classic()+
#   geom_hline(yintercept = 0, col="black", linetype="dashed") +
#   geom_errorbar(width=.2, size=0.5, stat="identity",position=position_dodge(width = 0.6)) +
#   geom_point(size=2, position=position_dodge(width = 0.6)) +
#   #GGPoptions +
#   ylab("Abundance (% difference from primary)") + 
#   theme(panel.spacing.y =  unit(0, "lines")) +
#   theme(axis.text.x = element_text(angle = 60, hjust=1, size = 12)) +
#   scale_discrete_manual("colour",
#                         values = c('black', 'red4', 'orange', 'darkgreen'),
#                         breaks=c('No intensity in model', 'Intense use', 'Light use', 'Minimal use')) +
#   ggtitle("Effects of models on total abundance (with and without land-use intensity)")

ggarrange(
## plotting the effects of all considered land uses on abundance
ggplot(abundance_predictions_all_SV %>% 
         filter(Use_intensity=='No intensity in model'),
       aes(LandUse , Median, ymin = Lower, ymax = Upper, group=Use_intensity, col=Use_intensity)) +
  xlab("") +
  theme_classic()+
  geom_hline(yintercept = 0, col="black", linetype="dashed") +
  geom_errorbar(width=.2, size=0.5, stat="identity",position=position_dodge(width = 0.6)) +
  geom_point(size=2, position=position_dodge(width = 0.6)) +
  #GGPoptions +
  ylab("Abundance (% difference from primary)") + 
  theme(panel.spacing.y =  unit(0, "lines")) +
  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 12)) +
  scale_discrete_manual("colour",
                        values = c('black', 'red4', 'orange', 'darkgreen'),
                        breaks=c('No intensity in model', 'Intense use', 'Light use', 'Minimal use')) +
  ggtitle("Effects of models on total abundance (with and without land-use intensity)")

,

ggplot(similarity_predictions,
       aes(contrast , Median, ymin = Lower, ymax = Upper)) +
  xlab("") +
  theme_classic()+
  geom_hline(yintercept = 0, col="black", linetype="dashed") +
  geom_errorbar(width=.2, size=0.5, stat="identity",position=position_dodge(width = 0.6)) +
  geom_point(size=2, position=position_dodge(width = 0.6)) +
  #GGPoptions +
  ylab("Abundance (% difference from primary)") + 
  theme(panel.spacing.y =  unit(0, "lines")) +
  theme(axis.text.x = element_text(angle = 60, hjust=1, size = 12)) +
  ggtitle("(b) Model on similarity") + scale_x_discrete( 
                    labels=c("PV-PV","PV-MSV","PV-ISV","PV-YSV",
                             "PV-PF", "PV-OP", "PV-Cr"))

)
```

# 4. MAPPING RESULTS: BII for mill areas

Here, we predict BII for the World Cover maps corrected for oil palm plantations, for the years 2020 and 2021.

/danone_matched_mills.csv')

MillsFerrero \<- read.csv('../../data/supply_shed_data/mills_5\_companies/ferrero_matched_mills.csv')

MillsNestle \<- read.csv('../../data/supply_shed_data/mills_5\_companies/nestle_matched_mills.csv')

MillsPepsico \<- read.csv('../../data/supply_shed_data/mills_5\_companies/pepsico_matched_mills.csv')

MillsWilmar \<- read.csv('../../data/supply_shed_data/mills_5\_companies/danone_matched_mills.csv')

```{r}
#| output: false 
#| echo: true 

## ESA world cover map
## https://protect.checkpoint.com/v2/r02/___https://developers.google.com/earth-engine/datasets/catalog/ESA_WorldCover_v200___.YzJlOmlpYXNhOmM6bzo0NzFlNmM2NWViNThhN2EyNTI1Nzk5ZGZhZjBmNzI5Zjo3OmI2NmU6ZTIyZGM3NjI1N2EzODlhMDNlMWM0YTZlYTgxYjU0ZDE4MWQxOWJlY2MwM2IxZWQyZGUzZDZjNzMzZTc4ZDhkNDp0OlQ6Tg

## loading mills for each company
MillsDanone <- read.csv('../../data/supply_shed_data/mills_5_companies/danone_matched_mills.csv')
MillsFerrero <- read.csv('../../data/supply_shed_data/mills_5_companies/ferrero_matched_mills.csv')
MillsNestle <- read.csv('../../data/supply_shed_data/mills_5_companies/nestle_matched_mills.csv')
MillsPepsico <- read.csv('../../data/supply_shed_data/mills_5_companies/pepsico_matched_mills.csv')
MillsWilmar <- read.csv('../../data/supply_shed_data/mills_5_companies/danone_matched_mills.csv')

## unique mills across the 5 companies
UniqueMills <- 
unique(c(MillsDanone$UML.ID,
  MillsFerrero$UML.ID, 
  MillsNestle$UML.ID, 
  MillsPepsico$UML.ID, 
  MillsWilmar$UML.ID))
length(UniqueMills)

## create a dataset with the mill locations
MillLocations <- 
unique(rbind(MillsDanone[,c('UML.ID' ,'Latitude', 'Longitude')],
  MillsFerrero[,c('UML.ID' ,'Latitude', 'Longitude')], 
  MillsNestle[,c('UML.ID' ,'Latitude', 'Longitude')], 
  MillsPepsico[,c('UML.ID' ,'Latitude', 'Longitude')], 
  MillsWilmar[,c('UML.ID' ,'Latitude', 'Longitude')]))


Run <- FALSE

if(Run){
  
# ### Test runs
# ## for mills 1 to 10
#   print(Sys.time())
#   T1 <- Sys.time()
#   UML_test <- Predict_BII(
#     UML_data = MillLocations[1:2, ],
#     Buffer_mill = 25000,
#     WC_layer = World_Cover_2021_crop,
#     Year = 2021,
#     FI_layer = Forest_integrity_crop,
#     Aggregate_Factor = 10,
#     abundance_model = abundance_model_SV_treecover,
#     similarity_model = similarity_model_tree_cover,
#     Path_raster_out = '../../results/2_biodiversity_metrics_site_level/BII_rasters_UML/02_04_2026/'
#   )
#   print(Sys.time())
#   T2 <- Sys.time()
#   print(T1 - T2)
 

## running on all the mills for the 5 different companies
## (e.g., correspond to the most recent 2024-H2 report from Danone)
## for 2021
print(Sys.time())
T1 <- Sys.time()
Mills5_2021 <- Predict_BII(
  UML_data = MillLocations,
  Buffer_mill = 25000,
  WC_layer = World_Cover_2021_crop,
  Year = 2021,
  FI_layer = Forest_integrity_crop,
  Aggregate_Factor = 10,
  abundance_model = abundance_model_SV_treecover,
  similarity_model = similarity_model_tree_cover,
  Path_raster_out = '../../results/2_biodiversity_metrics_site_level/BII_rasters_UML/09_04_2026/'
)
print(Sys.time())
T2 <- Sys.time()
print(T1-T2) 

write.csv(Mills5_2021, '../../results/2_biodiversity_metrics_site_level/BII_mills_tables/April_2026/5_companies_mills_2021.csv', row.names=FALSE)

## for 2020
print(Sys.time())
T1 <- Sys.time()
Mills5_2020 <- Predict_BII(
  UML_data = MillLocations,
  Buffer_mill = 25000,
  WC_layer = World_Cover_2020_crop,
  Year = 2020,
  FI_layer = Forest_integrity_crop,
  Aggregate_Factor = 10,
  abundance_model = abundance_model_SV_treecover,
  similarity_model = similarity_model_tree_cover,
  Path_raster_out = '../../results/2_biodiversity_metrics_site_level/BII_rasters_UML/09_04_2026/'
)
print(Sys.time())
T2 <- Sys.time()
print(T1-T2) 

write.csv(Mills5_2020, '../../results/2_biodiversity_metrics_site_level/BII_mills_tables/April_2026/5_companies_mills_2020.csv', row.names=FALSE)

}

```

Analysing the results: plotting Median and Sum BII distributions for each of the 5 compamies.

```{r}
#| output: false 
#| echo: true 

## consolidating the BII data
BII_2021 <- read.csv('../../results/2_biodiversity_metrics_site_level/BII_mills_tables/April_2026/5_companies_mills_2021.csv') %>% 
  mutate(Year=2020)
BII_2020 <-  read.csv('../../results/2_biodiversity_metrics_site_level/BII_mills_tables/April_2026/5_companies_mills_2020.csv') %>% 
  mutate(Year=2021)

## adding relevant BII values for each company
LJf <- function(Mills_company, BII2020, BII2021) {
  BII_2020 <- left_join(Mills_company, BII2020)
  BII_2021 <- left_join(Mills_company, BII2021)
  return(rbind(BII_2020, BII_2021))
}

BII_Danone <- LJf(MillsDanone, BII_2020, BII_2021)
BII_Pepsico <- LJf(MillsPepsico, BII_2020, BII_2021)
BII_Ferrero <- LJf(MillsFerrero, BII_2020, BII_2021)
BII_Nestle <- LJf(MillsNestle, BII_2020, BII_2021)
BII_Wilmar <- LJf(MillsWilmar, BII_2020, BII_2021)

BII_all <- rbind(BII_Danone %>% 
                   dplyr::select(UML.ID, Latitude, Longitude, Year, Median_BII, Sum_BII) %>% 
                   mutate(Company='Danone'), 
                 BII_Pepsico %>% 
                   dplyr::select(UML.ID, Latitude, Longitude, Year, Median_BII, Sum_BII) %>% 
                   mutate(Company='Pepsico'), 
                 BII_Ferrero %>% 
                   dplyr::select(UML.ID, Latitude, Longitude, Year, Median_BII, Sum_BII) %>% 
                   mutate(Company='Ferrero'), 
                 BII_Nestle %>% 
                   dplyr::select(UML.ID, Latitude, Longitude, Year, Median_BII, Sum_BII) %>% 
                   mutate(Company='Nestle'), 
                 BII_Wilmar %>% 
                   dplyr::select(UML.ID, Latitude, Longitude, Year, Median_BII, Sum_BII) %>% 
                   mutate(Company='Wilmar')
                 )

write.csv(BII_all, '../../results/2_biodiversity_metrics_site_level/BII_mills_tables/April_2026/5_companies_BII.csv', row.names=FALSE)
```

```{r}
#| output: true 
#| echo: true 

## plotting distributions of BII for each company
ggplot(BII_all, aes(Median_BII, col=Company, fill=Company)) + 
  geom_density(alpha=0.1, size=0.6) +
  facet_wrap(~Year) +
  theme_bw()

ggplot(BII_all, aes(Sum_BII, col=Company, fill=Company)) + 
  geom_density(alpha=0.1, size=0.6) +
  facet_wrap(~Year) +
  theme_bw()
```

```{r}
#| output: true 
#| echo: true 

# ## changes from 2020 to 2021 in median BII
# Medians2020 <- Set$Median_BII[Set$Year==2020]
# Medians2021 <- Set$Median_BII[Set$Year==2021]
# MediansYears <- as.data.frame(cbind(Medians2020, Medians2021))
# MediansYears$delta <-  MediansYears$Medians2020 - MediansYears$Medians2021    
# MediansYears$UML_ID <-  Set$UML.ID[Set$Year==2020]
# 
# ## changes from 2020 to 2021 in sum BII
# Sum2020 <- Set$Sum_BII[Set$Year==2020]
# Sum2021 <- Set$Sum_BII[Set$Year==2021]
# SumYears <- as.data.frame(cbind(Sum2020, Sum2021))
# SumYears$delta <-  SumYears$Sum2020 - SumYears$Sum2021    
# SumYears$UML_ID <-  Set$UML.ID[Set$Year==2020]
# 
# ggarrange(
# ggplot(MediansYears, aes(Medians2020*100, Medians2021*100)) + 
#   geom_point() + 
#   theme_bw() + 
#   geom_abline(slope = 1, intercept = 0, col='red', lty='dashed') +
# ggtitle('Median BBI across mills (2020 versus 2021)')  +
#   xlab('Median BII - 2020') + ylab( xlab('Median BII - 2021'))
# ,
# ggplot(SumYears, aes(Sum2020*100, Sum2021*100)) + 
#   geom_point() + 
#   theme_bw() + 
#   geom_abline(slope = 1, intercept = 0, col='red', lty='dashed') +
# ggtitle('Sum BBI across mills (2020 versus 2021)')  +
#   xlab('Sum BII - 2020') + ylab( xlab('Sum BII - 2021'))
# )

```

```{r}
#| output: true 
#| echo: true 

## plotting example
# Example2020 <- rast("../../results/2_biodiversity_metrics_site_level/BII_rasters_UML/13_02_2026/2020/PO1000000169.tif")
# 
# Example2021 <- rast("../../results/2_biodiversity_metrics_site_level/BII_rasters_UML/13_02_2026/2021/PO1000000169.tif")
# 
# plot(Example2020*100, main='BII (% difference from primary vegetation baseline) \n- Mill ID: PO1000000052')
# 
# plot(Example2021*100, main='BII (% difference from primary vegetation baseline) \n- Mill ID: PO1000000052')

Example2020 <- rast("../../results/2_biodiversity_metrics_site_level/BII_rasters_UML/09_04_2026/2020/PO1000000052.tif")
Example2020_2 <- rast("../../results/2_biodiversity_metrics_site_level/BII_rasters_UML/09_04_2026/2020/PO1000003715.tif")

Example2021 <- rast("../../results/2_biodiversity_metrics_site_level/BII_rasters_UML/09_04_2026/2021/PO1000003951.tif")
Example2021_2 <- rast("../../results/2_biodiversity_metrics_site_level/BII_rasters_UML/09_04_2026/2021/PO1000001252.tif")


par(mfrow=c(1,2))
plot(Example2020_2*100, main='BII (% difference from primary vegetation baseline) \n- Mill ID: PO1000000052; 2020', cex.main=0.8)
plot(Example2021_2*100, main='BII (% difference from primary vegetation baseline) \n- Mill ID: PO1000003715; 2020', cex.main=0.8)
# plot(Example2021*100, main='BII (% difference from primary vegetation baseline) \n- Mill ID: PO1000003951; 2021', cex.main=0.8)
# plot(Example2021_2*100, main='BII (% difference from primary vegetation baseline) \n- Mill ID: PO1000001252; 2021', cex.main=0.8)
```

# 4. CONCLUSIONS

We found significant biodiversity declines in oil-palm plantations compared to primary vegetation. Total abundance declined by an average 15%, and similarity by an average 43%. We provide BII for mills for 5 companies in Indonesia and Malaysia, considering a 25km buffer area around the mills.

Challenges to think further about: - how to better align PREDICTS categories with EO-products on land use and land use intensity, including: --\> how to get proxies for land use intensity --\> how to better represent primary and secondary vegetation in world cover maps.

------------------------------------------------------------------------

Time required to compile the document:

```{r}
#| output: true 
#| echo: true
TimeEnd <- Sys.time()
print(TimeEnd-TimeStart)

```
